---
title: "Final Exam on Scorpion Behavior"
author: "Put your name here!"
date: "put the date here"
output: word_document
---

```{r include=FALSE,echo=FALSE}
require(knitr)
require(mosaic)
require(UsingR)
require(tigerstats)

```

```{r}
url2="scorpion_reaction_rolls.csv"
scorp_data2=read.csv(url2,header=T)
names(scorp_data2)
attach(scorp_data2)
rol=table(roll,Behavior)
rol
chisqtestGC(rol)
sze=table(size,Behavior)
chisqtestGC(sze)
```


```{r}
contTable <- xtabs(~ Behavior + size, data = scorp_data2)
barchart(contTable,
	auto.key = list(
		space = "top",
		title = "size",
		cex.title = 1,
		columns = 1),
	stack = FALSE , horizontal = FALSE)
```



# Introduction

Investigate whether the odors from a predator might influence the defensive behavior of Hadrurus arizonensis, and if the scorpion size plays a role in the type of response.

we see this as two questions:

Does presence of odor change the expected outcome on our measurements?


#### what was asked for

Investigate whether the odors from a predator might influence the defensive behavior of Hadrurus arizonensis, and if the scorpion size plays a role in the type of response.

we see this as two questions

##### does presence of odor change the expected outcome on our measurements?

<for continuous measurements we could ask for center differences (means or medians)

for catagorical explanitory variables (such as count of type of response)
a chi-squared test would be chosen.>

##### does size of the scorpion change the expected outcome on our measurements?

<for continuous measurements we could ask for center differences (means or medians)

for catagorical explanitory variables (such as count of type of response)
a chi-squared test would be chosen.>


# Methods for presence vs absence of odor

<Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.>

#Results

<Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.>

## Descriptive Results

In order to understand the comparison between the scorpions when odor is a factor to explain when they would attack or become more aggrevated. We must create a plot in order to compare the reaction time for the scorpions when they feel threatened by a predator and when they do not feel threatened by a predator. We must also produce a t.test in order to show the comparison between the control group as well as the experimental group so that we may see if they have a significant comparison. We will begin by showing our head(rol), which describe our results from our Scorpion Analysis.

```{r}
head(rol)
```


### Graphical Descriptive Results

```{r}
contTable <- xtabs(~ Behavior + roll, data = scorp_data2)
barchart(contTable,
    auto.key = list(
        space = "top",
        title = "roll",
        cex.title = 1,
        columns = 1),
    stack = FALSE , horizontal = FALSE)
```

The bar graph clearly shows that the scorpions are much more likely to have a wet sting when a threat is present, than in the control group when the threat is not present.

This plot shows that when we factor the reaction time for both the control and experimental group the comparison has a positive correlation which would indicate that the reaction time for the experiemental group has a direct correlation to the reaction time for the control group.

### Numerical Descriptive Results

```{r}
attach(scorp_data2)
```

```{r}
rol=table(roll,Behavior)
rowPerc(rol)
```

During the rowPerc test it clearly shows that the scorpions are more likely to have a wet sting when a threat is present as opposed to when there is no threat present. About 70% of the scorpions will have a wet sting when the threat is present; while those who did not have a threat present only about 7% had a wet sting.

## Inferential Results

```{r}
chisqtestGC(rol,simulate.p.value = TRUE ,graph=TRUE)
```

The chisqtest shows that we would reject the null hypothesis based on the fact that the wet sting was used more frequently with the threat prescence in comparison to when there is no threat present. The pvalue is significantly low which indicates that there is a difference between the two variables compared.

<State hypothesis clearly. Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc. Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question>

# Methods for different sizes (large vs small)

<Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.>

#Results

<Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.>

## Descriptive Results

### Graphical Descriptive Results

```{r}
contTable <- xtabs(~ Behavior + size, data = scorp_data2)
barchart(contTable,
	auto.key = list(
		space = "top",
		title = "size",
		cex.title = 1,
		columns = 1),
	stack = FALSE , horizontal = FALSE)
```
The small scorpions have more of a reaction of the dry sting , while the larger scorpions have more of a wet sting reaction. There is no change in the no sting catergory, the large scorpions created a no sting reaction.


### Numerical Descriptive Results

```{r}
rol=table(size,Behavior)
rowPerc(rol)
```

The outcome of our table did show that there was fundamental differences between the large and the small scorpions, being that the expected outcome shows that the adult scorpion shows a slight insignificant (.22%) higher percentage of wet stings. The biggest difference is in the dry sting category with the small scorpions giving you a significant higher percentage than any other of the categories represented on our table. 
## Inferential Results


```{r}
chisqtestGC(rol,simulate.p.value = TRUE ,graph=TRUE)
```

Based on my expected null behavior my chi square statistic is five times greater then my degrees of freedom, and my p-value is almost zero. We must reject the Null Hypothesis based on the P Value is 0.001.  The possibilty of the behaviors being different are almost at 0 chance.

# Discussion and Conclusion for both set of research questions

<Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.>
